---
title: 'Bios 6301: Assignment 2'
author: "Abdurrahman Abdulhamid"
date: "`r Sys.Date()`"
output: html_document
---

#Section 1 Working with data
  
# 1. Load the data set into R and make it a data frame called `cancer.df`
```{r}
data <- read.csv('cancer.csv')
cancer.df <- read.csv("cancer.csv", header = TRUE, sep = ",")
nrow(cancer.df)
ncol(cancer.df)
```

# 2 Determine the number of rows and columns in the data frame.
```{r}
cancer.df <- read.csv("cancer.csv", header = TRUE, sep = ",")
nrow(cancer.df)
ncol(cancer.df)
```

# 3. Extract the names of the columns in `cancer.df`. (2)

    ```{r}
colnames(cancer.df)
```
# 4 Report the value of the 3000th row in column 6.

```{r}
v1 <- cancer.df[c(3000),c(6)]
v1
```
# 5 Report the contents of the 172nd row.
```{r}
v2 <- cancer.df[172,]
v2
```
# 6 Create a new column that is the incidence *rate* (per 100,000) for each row. The incidence rate is the`(number of cases)/(population at risk)`, which in this case means `(number of cases)/(population at risk) * 100,000`

```{r}
cancer.df1 <- transform(cancer.df, incidence_rate = `incidence`/`population`*100,000)
View(cancer.df1)
```

# 7 How many subgroups (rows) have a zero incidence rate?

```{r}
df<-data.frame(cancer.df1)
df
df3 <- data.frame(colSums(df==0))
View(df3)
nrow(df3)
v3 <-data.frame(df3)
v4 <- v3[c(9),c(1)]
#or
V5 <- v3[9,1]
```
# 8 Find the subgroup with the highest incidence rate.

```{r}
max(cancer.df1$incidence_rate)
```

# Section Two
**Data types**

# 1. Create the following vector: `x <- c("5","12","7")`. Which of the following commands will produce an error message? For each command, Either explain why they should be errors, or explain the non-erroneous result.

```{r}
x <- c("5","12","7")
#max(x)
#sort(x)
#sum(x)
#The first command max(x) extracted the highest value in the column vector but because of the invalid "type" character of argument it does not obey the command.#
#Sort(x) because of the invalid 'type' (character) of argument, the command sort the first two entries excluding the third one in ascending order instead of arranging all the entries in ascending order.#
#sum(x) is a command adding all the entries of the column vector, but here the vector x has invalid "type" (character) of argument.#
```
#  2. For the next two commands, either explain their results, or why they should produce errors.

```{r}
            y <- c("5",7,12)
            #y[2] + y[3]
# The first command neither create a column vector nor display any informatio due to invalid 'type'.
#For the second command, there is an error in y[2] + y[3] : non-numeric argument to binary operator#
```
#3. For the next two commands, either explain their results, or why they should produce errors.
```{r}
z <- data.frame(z1="5",z2=7,z3=12)
z
z[1,2] + z[1,3] 
# z if displayed will be 5, 7, 12 so z[1,2] will identify entry in the first row second column (7) and add it to the one in first row third column (12) which gives 19.
```
# Section 3 **Data structures**

#Give R expressions that return the following matrices and vectors (*i.e.* do not construct them manually).

```{r}
x <- seq(1,7)
y <- seq(8,1,-1) 
z<-list (c(seq(1,7),seq(8,1,-1)))
#Section 3 Question 1
print(z)
#Section 3 Question 2
rep(1:5, 1:5)
mt <- diag(1, 3, 3) # [upper.tri] <- 0)
mt[upper.tri(mt)] <- 0
mt[lower.tri(mt)] <- 0
#Section 3 Question 3
print(mt)
x1 <- c(1:4)
x2 <- x1^2
x2
x3 <- x1^3
x4 <- x1^4
x5 <- x1^5
# creating matrix
n <- rbind(x1, x2, x3, x4, x5)
#Section 3 Question 4
# print matrix
print(n)
```
    1. $(1,2,3,4,5,6,7,8,7,6,5,4,3,2,1)$
    
    2. $(1,2,2,3,3,3,4,4,4,4,5,5,5,5,5)$

    3. $\begin{pmatrix}
      0 & 1 & 1  \\
      1 & 0 & 1  \\
      1 & 1 & 0  \\
    \end{pmatrix}$

    4. $\begin{pmatrix}
      1 & 2 & 3 & 4 \\
      1 & 4 & 9 & 16 \\
      1 & 8 & 27 & 64  \\
      1 & 16 & 81 & 256 \\
      1 & 32 & 243 & 1024  \\
    \end{pmatrix}$

4. **Basic programming** (10 points)

```{r}



```

    1. Let $h(x,n)=1+x+x^2+\ldots+x^n = \sum_{i=0}^n x^i$. Write an R program to calculate $h(x,n)$ using a `for` loop. As an example, use `x = 5` and `n = 2`. (5 points)

    1. If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23. Write an R program to perform the following calculations. (5 points)

        1. Find the sum of all the multiples of 3 or 5 below 1,000. (3, [euler1])

        1. Find the sum of all the multiples of 4 or 7 below 1,000,000. (2)

    1. Each new term in the Fibonacci sequence is generated by adding the previous two terms. By starting with 1 and 2, the first 10 terms will be $(1, 2, 3, 5, 8, 13, 21, 34, 55, 89)$. Write an R program to calculate the sum of the first 15 even-valued terms. (5 bonus points, [euler2])

Some problems taken or inspired by projecteuler.

[euler1]: https://projecteuler.net/problem=1
[euler2]: https://projecteuler.net/problem=2

 1. **Working with data** In the `datasets` folder on the course GitHub repo, you will find a file called `cancer.csv`, which is a dataset in comma-separated values (csv) format. This is a large cancer incidence dataset that summarizes the incidence of different cancers for various subgroups. (18 points)
 
    1. Load the data set into R and make it a data frame called `cancer.df`. (2 points)

    2. Determine the number of rows and columns in the data frame. (2)

    3. Extract the names of the columns in `cancer.df`. (2)

    4. Report the value of the 3000th row in column 6. (2)

    5. Report the contents of the 172nd row. (2)

    6. Create a new column that is the incidence *rate* (per 100,000) for each row. The incidence rate is the `(number of cases)/(population at risk)`, which in this case means `(number of cases)/(population at risk) * 100,000`. (3)

    7. How many subgroups (rows) have a zero incidence rate? (2)

    8. Find the subgroup with the highest incidence rate.(3)

2. **Data types** (10 points)

    1. Create the following vector: `x <- c("5","12","7")`. Which of the following commands will produce an error message? For each command, Either explain why they should be errors, or explain the non-erroneous result. (4 points)

            max(x)
            sort(x)
            sum(x)

    2. For the next two commands, either explain their results, or why they should produce errors. (3 points)

            y <- c("5",7,12)
            y[2] + y[3]

    3. For the next two commands, either explain their results, or why they should produce errors. (3 points)

            z <- data.frame(z1="5",z2=7,z3=12)
            z[1,2] + z[1,3]

3. **Data structures** Give R expressions that return the following matrices and vectors (*i.e.* do not construct them manually). (3 points each, 12 total)

    1. $(1,2,3,4,5,6,7,8,7,6,5,4,3,2,1)$
    
    2. $(1,2,2,3,3,3,4,4,4,4,5,5,5,5,5)$

    3. $\begin{pmatrix}
      0 & 1 & 1  \\
      1 & 0 & 1  \\
      1 & 1 & 0  \\
    \end{pmatrix}$

    4. $\begin{pmatrix}
      1 & 2 & 3 & 4 \\
      1 & 4 & 9 & 16 \\
      1 & 8 & 27 & 64  \\
      1 & 16 & 81 & 256 \\
      1 & 32 & 243 & 1024  \\
    \end{pmatrix}$

4. **Basic programming** (10 points)

    1. Let $h(x,n)=1+x+x^2+\ldots+x^n = \sum_{i=0}^n x^i$. Write an R program to calculate $h(x,n)$ using a `for` loop. As an example, use `x = 5` and `n = 2`. (5 points)

    1. If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23. Write an R program to perform the following calculations. (5 points)

        1. Find the sum of all the multiples of 3 or 5 below 1,000. (3, [euler1])

        1. Find the sum of all the multiples of 4 or 7 below 1,000,000. (2)

    1. Each new term in the Fibonacci sequence is generated by adding the previous two terms. By starting with 1 and 2, the first 10 terms will be $(1, 2, 3, 5, 8, 13, 21, 34, 55, 89)$. Write an R program to calculate the sum of the first 15 even-valued terms. (5 bonus points, [euler2])

Some problems taken or inspired by projecteuler.

[euler1]: https://projecteuler.net/problem=1
[euler2]: https://projecteuler.net/problem=2